Paper

### 2018-03-06 Paper.Rmd
adding to pipeline

ggsave('plots/performance/no_lgd_perf.png', plot=mp$plot, height=5, width=5, units='in', dpi=30)

@@

droc = multi.roc.plot(list(roc), colors=brewer.pal(3,'Greys')[3]) + labs(title='Discovery cohort ROC')
droc
ggsave(filename='plots/discovery_roc.png', plot=droc, width=6, height=4, units='in', dpi=300)

@@

@@

ggplot(m, aes(Patient,  value*100, group=variable, fill=variable)) + geom_bar(stat='identity', position=position_stack()) +
  geom_text(aes(label=n.samples, y=101)) + scale_fill_manual(values=riskCols) +
  facet_grid(~Status, scales='free_x') + labs(y='Percent of samples', title='Validation patient set') + plot.theme
```

@@

@@

cols = unique(colnames(back), colnames(back.v))
cols = cols[-grep('quants|pred\\.ratio|se\\(Risk\\)', cols)]
allB = rbind.data.frame(back[,cols], back.v[,cols])
allB$p53.Status = factor(allB$p53.Status, levels=c(0,1), ordered = T)

@@

second = subset(rules,  t2 != 'Endpoint') %>% 
  group_by(Patient, Status) %>% 
  summarise(  max(t2), 'rule'=rule[t2 == max(t2)]  )
table(subset(second, Status == 'P')$rule)
table(subset(second, Status == 'NP')$rule)

@@

ggplot(rrPath, aes(Status, mean.RR)) + #scale_color_gradientn(colors=myPal, limits=c(min(rrPath[,3:5]), max(rrPath[,3:5])),  name='') +
  facet_wrap(~Pathology, ncol=1) + 
  geom_point() +  geom_errorbar(aes(ymin=ci.min, ymax=ci.max), width=0.1) + 
  geom_text(aes(y=ci.min, label=round(ci.min,2)), vjust=2) + geom_text(aes(y=ci.max, label=round(ci.max,2)), vjust=2) +
  coord_flip() + plot.theme + labs(y='Mean RR', x='')

@@

as.data.frame(subset(back, Status == 'P' & OR < 2 & OR > -2, select=c('Status','Patient','Pathology','months.before.final','OR')))

hist(subset(back, Pathology %in% c('BE'))$OR)

@@

ggplot(melt(loopreds[c('Min','Max')]), aes(value)) + facet_wrap(~variable, ncol=2) +
  geom_histogram(aes(fill=..x..), breaks=cuts, show.legend = F) + 
    scale_fill_gradientn(colors = myPal,  name='') + 
    plot.theme + theme(legend.position = 'none') + 
    labs(y='n Samples', x='Absolute Risk',
         title='5-year adjusted risk', subtitle='Adjusted for upper & lower bound: 1-3.5% risk')

>>>>

ggsave('plots/performance/no_lgd_perf.png', plot=mp$plot, height=5, width=5, units='in', dpi=300)

@@

droc = multi.roc.plot(list(roc), colors=brewer.pal(3,'Greys')[3]) + labs(title='Discovery cohort ROC')
droc
ggsave(filename='plots/auc/discovery_roc.png', plot=droc, width=6, height=4, units='in', dpi=300)

@@

df = subset(globalCNdf, Patient == 20 & seg.type != 'arms')
#df = subset(globalCNdf, Patient == 51 & seg.type != 'arms')

# early/late samples
#df = subset(globalCNdf, Patient == pt)
df = merge(globalCNdf, back[,c('Samplename','Endoscopy.Year', 'ogj', 'months.before.final', 'OR','Prediction')], by.x='variable', by.y='Samplename')
df = arrange(df, chr, start, Endoscopy.Year)

## count arms differently than segs??
df2 = df %>% group_by(Status, Patient, Endoscopy.Year) %>% summarise( 
  'max.Pred'=max(Prediction),
  'max.RR'=max(OR),
  'max.Path'=max(Pathology),
  'CNA'=length(which(seg.type != 'arms' & cn %in% c('loss','gain'))),
  'arms'=length(which(seg.type == 'arms' & cn %in% c('loss','gain'))),
  'ratio'=length(which(seg.type != 'arms' & cn %in% c('loss','gain')))/length(cn)
  )
#as.data.frame(df2)
#df2 = subset(df2, max.Path %in% c('BE','ID','LGD'))

df2 = df2 %>% group_by(Status, Patient) %>% mutate(
  'early'= Endoscopy.Year == min(Endoscopy.Year),
  'late' = Endoscopy.Year == max(Endoscopy.Year)
)
df2 = subset(df2, Patient %in% names(which(table(df2$Patient) > 1)))

el = subset(df2, early | late)
el2 = el %>% group_by(Status, Patient) %>% summarise( diff(arms), diff(CNA) )
wtA = round(wilcox.test(subset(el2, Status == 'Progressor')$`diff(arms)`, subset(el2, Status == 'Non-Progressor')$`diff(arms)`)$p.value, 3)
if (wtA < 0.0001) wtA = '< 0.0001'
#wilcox.test(subset(el2, Status == 'Progressor')$`diff(CNA)`, subset(el2, Status == 'Non-Progressor')$`diff(CNA)`)

p = ggplot(subset(df2, early|late), aes(Endoscopy.Year, arms, color=Patient)) + facet_grid(~Status) +
  geom_point() +  geom_line(data=subset(df2, early | late)) + theme(legend.position = 'none', text=element_text(size=11)) + 
  plot.theme + labs(x='Endoscopy Year', y='CN arm values', title='Initial vs final endoscopy', subtitle=paste('p', wtA))
p
ggsave(filename='plots/status_hetero.png', plot=p, width=8, height=6, units='in', dpi=300)

wtA = round(wilcox.test(subset(df2, early &  Status == 'Progressor')$arms, subset(df2, early & Status == 'Non-Progressor')$arms)$p.value, 3)
wtCN = round(wilcox.test(subset(df2, early & Status == 'Progressor')$CNA, subset(df2, early & Status == 'Non-Progressor')$CNA)$p.value, 2)

m = melt(subset(df2, early), measure.vars=c('CNA','arms'), id.vars=c('Status'))
p = ggplot(m, aes(Status, value, fill=Status)) + facet_wrap(~variable, scales='free_y') + 
  geom_boxplot(outlier.colour = NA) + geom_jitter(width=0.2) + plot.theme + theme(legend.position = 'none', text=element_text(size=11)) +
  labs(y='', x='', title='CN values, initial endoscopy', subtitle=paste('Arms p=',wtA, '   CNA p=',wtCN, sep=''))
p
ggsave(filename='plots/hetero_initial.png', plot=p, width=6, height=4, units='in', dpi=300)

ggplot(el2, aes(Status,`diff(arms)`)) + geom_boxplot() + geom_jitter(width=0.2)
ggplot(el2, aes(Status,`diff(CNA)`)) + geom_boxplot() + geom_jitter(width=0.2)

@@

ggplot(m, aes(Patient,  value*100, group=variable, fill=variable)) + geom_bar(stat='identity', position=position_stack()) +
  geom_text(aes(label=n.samples, y=101)) + scale_fill_manual(values=riskCols) +
  facet_grid(~Status, scales='free_x') + labs(y='Percent of samples', title='Validation patient set') + plot.theme

back.v$Prediction

table(subset(back.v, Status == 'P')$Prediction > 0.6)
table(subset(back.v, Status == 'NP')$Prediction < 0.3)

@@

## Downsampled Datasets

```{r echo=F, eval=T}
dirA = '~/Data/Ellie/Analysis/downsampled5G_BEAdjacent'
dirB = '~/Data/Ellie/Analysis/downsampled5G_NDBE'

file = paste(cache.dir, 'model_data.Rdata', sep='/')
if (!file.exists(file))
  stop(paste("Missing data file", file))
load(file, verbose=T)
file = paste(cache.dir, 'all.pt.alpha.Rdata', sep='/')
if (!file.exists(file))
  stop(paste("Missing data file", file))
load(file, verbose=T)
rm(dysplasia.df,coefs,plots,labels)

select.alpha = '0.9'
fitV = models[[select.alpha]]
lambda.opt = performance.at.1se[[select.alpha]][, 'lambda']

predA = list.files(dirA, 'predictions.Rdata', full.names = T)
predB = list.files(dirB, 'predictions.Rdata', full.names = T)

files = c()
if (length(predA) < 1)
  files = c(list.files(dirA, pattern='raw', full.names=T))
if (length(predB) < 1)
  files = c(files, list.files(dirB, pattern='raw', full.names=T))

if (length(files) > 0) {
  sampleNames = sub('_raw.txt', '', basename(files))
  dspred = data.frame(matrix(ncol=2, nrow=length(sampleNames), dimnames=list(sampleNames,c('Prob','RR'))))

  for (n in 1:length(files)) {
    print(n)
    print(sampleNames[n])

    rs = read.table(files[n], sep='\t',header=T)
    if (ncol(rs) < 4)
      stop("File needs to contain 4 columns: chr, start, end, sample_value")

    segs <- tile.segmented.data(rs, size=5e6, chr.info=chrlen)
    segM = as.matrix(segs[,-c(1:3)])
    rownames(segM) = paste(segs$chr, ':', segs$start, '-', segs$end, sep='')
    segM = t(segM)
    for (i in 1:ncol(segM)) 
      segM[,i] = unit.var(segM[,i], z.mean[i], z.sd[i])

    arms <- tile.segmented.data(rs, size='arms', chr.info=chrlen)
    armsM = as.matrix(arms[,-c(1:3)])
    rownames(armsM) = paste(arms$chr, ':', arms$start, '-', arms$end, sep='')
    armsM = t(armsM)
    for (i in 1:ncol(armsM)) 
      armsM[,i] = unit.var(armsM[,i], z.arms.mean[i], z.arms.sd[i])

    nrow(armsM) == nrow(segM)

    cx.score = score.cx(segM,1)

    mergedDf = subtract.arms(segM, armsM)
    mergedDf = cbind(mergedDf, 'cx' = unit.var(cx.score, mn.cx, sd.cx))
    dim(mergedDf)

    # Predict function giving me difficulty when I have only a single sample, this ensures the dimensions are the same
    sparsed_test_data <- Matrix(data=0, nrow=nrow(mergedDf),  ncol=ncol(mergedDf),
                                dimnames=list(rownames(mergedDf),colnames(mergedDf)), sparse=T)
    for(col in colnames(mergedDf)) sparsed_test_data[,col] = mergedDf[,col]

    preds = predict(fitV, newx=sparsed_test_data, s=lambda.opt, type='response')
    RR = predict(fitV, newx=sparsed_test_data, s=lambda.opt, type='link')

    dspred[n,] = c(preds[,1],RR[,1])
    print(dspred[1:n,])
  }
} else {
  load(predB, verbose=T)
  dspredB = dspred

  load(predA, verbose=T)

  dspred = rbind(dspred, dspredB)
  dspred$Type = NA

  dspred[sub('_raw.*','',basename(list.files(dirA, pattern='raw', full.names=T))),'Type'] = 'AdjacentBE'
  dspred[sub('_raw.*','',basename(list.files(dirB, pattern='raw', full.names=T))),'Type'] = 'BEST2 NDBO'

}

ggplot(dspred, aes(Type, RR, fill=Type)) + geom_boxplot() + geom_jitter(width=0.2) +
  labs(x='', y='Relative Risk', title='Predictions on downsampled patients') + 
  theme(legend.position = 'none') + plot.theme

adj = subset(dspred, Type == 'AdjacentBE')
prg = back[,c('Prediction', 'OR', 'Status')]
prg$Status = ifelse(prg$Status == 'P', 'Progressor','Non-Progressor')

#  subset(back, Status == 'P', select=c('Prediction', 'OR', 'Status'))
colnames(prg) = colnames(adj)
m = rbind(adj,prg)
m$Type = factor(m$Type, levels=c('Non-Progressor','BEST2 NDBO','Progressor','AdjacentBE'), ordered = T)

pr = ggplot(m, aes(Type, RR, fill=Type)) + 
  geom_jitter(width=0.2) + geom_boxplot() + 
  labs(x='', y='RR', title='Predicted Relative Risk') + 
  theme(legend.position = 'none', text=element_text(size=11)) + plot.theme 
pr

ggsave('plots/adjacentBE.png', plot=pr, width=8, height=8,dpi=300)

#save(dspred, file=paste(dirB, 'predictions.Rdata', sep='/'))

@@

cols = intersect(colnames(back), colnames(back.v))
#cols = cols[-grep('quants|pred\\.ratio|se\\(Risk\\)', cols)]
allB = rbind.data.frame(back[,cols], back.v[,cols])
allB$p53.Status = factor(allB$p53.Status, levels=c(0,1), ordered = T)

@@

second = subset(rules,  t2 != 'Endpoint') %>% 
  group_by(Patient, Status) %>% 
  summarise(  max(t2), 'rule'=rule[t2 == max(t2)]  )

table(subset(second, Status == 'P')$rule)
table(subset(second, Status == 'NP')$rule)

path.rx<-function(Patient, brks, Pathology) {
  rules = as.data.frame(matrix(ncol=0, nrow=0))
  # Consecutive
  for (i in 1:length(brks)) {
    #risks = table(Pathology[i:(i+1)])
    rule = 'None'
    if ( length(which(grepl('HGD|IMC', Pathology[i:(i+1)]))) > 0  |
         length(which(grepl('LGD', Pathology[i:(i+1)]))) > 1 ) {
      rule = 1
    } else if ( length(which(grepl('LGD', Pathology[i:(i+1)]))) > 0 ) {
      rule = 2
    } else if ( length(which(grepl('BE|ID', Pathology[i:(i+1)]))) > 1  ) {
      rule = 3
    }
  rules = rbind(rules, cbind( as.character(brks[i]), as.character(brks[(i+1)]), rule))
  }  
  colnames(rules) = c('t1','t2', 'rule')
  rules$Patient = Patient
  rules
}

rules2 = do.call(rbind, lapply(unique(maxP$Patient), function(pt) {
  r = with(arrange(subset(maxP, Patient == pt), brks), path.rx(pt, brks,Pathology))
  r$Status = unique(subset(maxP, Patient == pt)$Status)
  r
}))
rules2$t1 = factor(rules2$t1, levels=levels(back$brks), ordered=T)
rules2$t2 = factor(rules2$t2, levels=levels(back$brks), ordered=T)

second = subset(rules2,  t2 != 'Endpoint') %>% 
  group_by(Patient, Status) %>% 
  summarise(  max(t2), 'rule'=rule[t2 == max(t2)]  )

table(subset(second, Status == 'P')$rule)
table(subset(second, Status == 'NP')$rule)

@@

rrp = ggplot(rrPath, aes(Status, mean.RR)) + #scale_color_gradientn(colors=myPal, limits=c(min(rrPath[,3:5]), max(rrPath[,3:5])),  name='') +
  facet_wrap(~Pathology, ncol=1) + 
  geom_point() +  geom_errorbar(aes(ymin=ci.min, ymax=ci.max), width=0.1) + 
  geom_text(aes(y=ci.min, label=round(ci.min,2)), vjust=2) + geom_text(aes(y=ci.max, label=round(ci.max,2)), vjust=2) +
  coord_flip() + plot.theme + labs(y='Mean RR', x='')
rrp

ggsave('plots/relrisk_path.png', plot=rrp, width=4, height=6,dpi=300)

@@

# as.data.frame(subset(back, Status == 'P' & OR < 2 & OR > -2, select=c('Status','Patient','Pathology','months.before.final','OR')))
# 
# hist(subset(back, Pathology %in% c('BE'))$OR)

@@

p = ggplot(melt(loopreds[c('Min','Max')]), aes(value)) + facet_wrap(~variable, ncol=2) +
  geom_histogram(aes(fill=..x..), breaks=cuts, show.legend = F) + 
    scale_fill_gradientn(colors = myPal,  name='') + 
    plot.theme + theme(legend.position = 'none') + 
    labs(y='n Samples', x='Absolute Risk',
         title='5-year adjusted risk', subtitle='Adjusted for upper & lower bound: 1-3.5% risk')
p

ggsave('plots/adj_RR.png', plot=p, width=6, height=6,dpi=300)
